---
title: "Exam 3"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(broom)
```

## Task 1: Faculty Salaries Plot

Let's look at the data structure before tidying.

```{r}
faculty_raw <- read.csv("FacultySalaries_1995.csv", header = FALSE, stringsAsFactors = FALSE)
faculty_raw
```
Oh yikes
```{r}
# Use the first row as column names
colnames(faculty_raw) <- faculty_raw[1, ]

# Remove first row from the data
faculty <- faculty_raw[-1, ]

# Take a look
head(faculty)
```

```{r}
names(faculty)
str(faculty)
```
Better!
```{r}
faculty_tidy <- faculty %>%
  select(State, Tier, AvgAssistProfSalary, AvgAssocProfSalary, AvgFullProfSalary) %>%
  pivot_longer(
    cols = c(AvgAssistProfSalary, AvgAssocProfSalary, AvgFullProfSalary),
    names_to = "Rank",
    values_to = "Salary"
  ) %>%
  mutate(
    Salary = as.numeric(Salary),
    Rank = recode(Rank,
                  AvgAssistProfSalary = "Assist",
                  AvgAssocProfSalary = "Assoc",
                  AvgFullProfSalary = "Full"),
    Rank = factor(Rank, levels = c("Assist", "Assoc", "Full")),
    Tier = factor(Tier, levels = c("I", "IIA", "IIB"))
  )

head(faculty_tidy)
str(faculty_tidy)
```
```{r}
faculty_tidy <- faculty_tidy %>%
  filter(!is.na(Tier))
```
This got rid of the NA


```{r}
ggplot(faculty_tidy, aes(x = Rank, y = Salary, fill = Rank)) +
  geom_boxplot() +
  facet_wrap(~ Tier) +
  labs(
    x = "Rank",
    y = "Salary",
    fill = "Rank"
  ) +
  theme_minimal()
```

## Task 2: ANOVA

```{r}
anova_model <- aov(Salary ~ State + Tier + Rank, data = faculty_tidy)

summary(anova_model)
```
State, Tier, and Rank all significantly affect salary (p < 0.001), indicating that salary varies across locations, institution types, and academic ranks.

## Task 3: Tidy Juniper Oils Data

```{r}
juniper <- read.csv("Juniper_Oils.csv")

head(juniper)
str(juniper)
```
```{r}
juniper_tidy <- juniper %>%
  select(YearsSinceBurn, contains(".")) %>%
  pivot_longer(
    cols = -YearsSinceBurn,
    names_to = "ChemicalID",
    values_to = "Concentration"
  )

head(juniper_tidy)
str(juniper_tidy)
```
## Task 4: Juniper Graph

```{r}
ggplot(juniper_tidy, aes(x = YearsSinceBurn, y = Concentration)) +
  geom_smooth(se = TRUE) +
  facet_wrap(~ ChemicalID, scales = "free_y") +
  labs(
    x = "YearsSinceBurn",
    y = "Concentration"
  ) +
  theme_minimal()
```

## Task 5: Linear Model

```{r}
model_results <- juniper_tidy %>%
  group_by(ChemicalID) %>%
  do(tidy(lm(Concentration ~ YearsSinceBurn, data = .))) %>%
  ungroup()
```

```{r}
significant_chemicals <- model_results %>%
  filter(term == "YearsSinceBurn", p.value < 0.05)

significant_chemicals
```

### ta daaaa